---
title: "Mapping US Vaccination Data in R"
author: "Michelle Chiu"
date: "7/14/2021"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required Packages

```{r packages, message = FALSE}
# uncomment below to install any missing packages as needed
# install.packages(c("readr", "tidyverse", "ggplot2", "viridis", "choroplethr", "maps"))

# load packages
library(readr)
library(tidyverse)
library(ggplot2)
library(viridis)
library(choroplethr)
library(choroplethrMaps)
library(maps)
```

### COVID-19 Vaccination Data

You can import the dataset **us_state_vaccinations.csv** from the data folder (downloaded on June 28, 2021) or download the most up-to-date version from [Data on COVID-19 vaccinations by Our World in Data](https://github.com/owid/covid-19-data).

#### Cleaning the data

Filter by 'date' and select the variables we will be plotting: 'location' and 'people_fully_vaccinated_per_hundred'

```{r data, message = FALSE}
# read covid-19 dataset into R and assign it to an object
covid_raw <- read_csv("data/us_state_vaccinations.csv")
head(covid_raw)

# filter by date and select relevant variables
today_covid <- 
  covid_raw %>% 
  filter(date=='2021-06-28') %>% 
  select(location, people_fully_vaccinated_per_hundred)
head(today_covid)
```

Read in the the US state map and assign it to an object **us_map**. The 'long' and 'lat' (longitude and latitude) columns will be used to generate our US state map.

```{r state_map, message = FALSE}
# map_data function from 'ggplot2' uses name of map provided by 'maps' package
# import us state map and assign it to an object
us_map <- map_data("state")
head(us_map) # peek at the us state map data
```

Next, we will tweak our **today_covid** dataframe to fit the format required for [state_choroplethr](https://arilamstein.com/documentation/choroplethr/reference/state_choropleth.html) function, then merge by 'region' to make our final dataframe **today_covid_map** that will be used in both [ggplot2](https://ggplot2.tidyverse.org/) and [choroplethr](https://arilamstein.com/packages/) map visuals.

```{r maps, message = FALSE}
# convert states to lowercase to match reference map
today_covid$location <- tolower(today_covid$location)

# rename column names and ny state to match reference map
colnames(today_covid) <- c('region','value')
today_covid$region <- 
  plyr::revalue(today_covid$region, c("new york state" = "new york"))

# merge vaccination data with map long/lat columns
today_covid_map <- merge(us_map, today_covid, by = "region")
```

### Mapping COVID-19 vaccination across the US

#### Using [ggplot2](https://ggplot2.tidyverse.org/)

```{r plot-1, warning = FALSE}
ggplot_map <- 
  ggplot(data = today_covid_map, 
         aes(x = long, y = lat)) + 
  geom_map(data = today_covid_map, map = us_map, 
           aes(fill = value, map_id = region)) +
  scale_fill_viridis(name = "fully vaccinated per 100 population",
                     option = "D") +
  theme_void() +
  labs(title = "State-by-state Vaccination Across the US",
       subtitle = "Data from the Centers for Disease Control and Prevention (June 28th, 2021)") +
  theme(text = element_text(size = 20),
        legend.title = element_text(size = 14),
        legend.position = 'right')

ggplot_map

```

#### Using [choroplethr](https://arilamstein.com/documentation/choroplethr/reference/)

```{r plot-2, warning = FALSE}
# note that state_choropleth parameters do not require us_map
choroplethr_map <- 
  state_choropleth(today_covid,
                   title = "State-by-state vaccination across the US",
                   legend = "Fully vaccinated per 100 population",
                   num_colors = 1)

choroplethr_map
# 5 quantiles, zoom in on vector of states
# state_choropleth(today_covid, num_colors = 5,
#                  zoom = c("new york", "pennsylvania", "new jersey"))
```

### Built-in Challenge: Choose your own adventure

We'll be using **state.x77** matrix built-in "state" data sets from Base R (R documentation [here](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/state)).
Check out solution.R script in the project repository for one (possible) set of solutions for the challenge.

```{r builtin-challenge, message = FALSE}
# Take the state.x77 data
# plot using choroplethr or ggplot or both
# `Frost`, mean number of days with minimum temperature below freezing (1931-1960)
# or one of the other vars you find interesting!

# This is the state.x77 dataset
head(state.x77)
?state.x77

# You will need to reformat/shape it to be used to make a plot!
# As a reminder, this is what the today_covid dataset looked like:
head(today_covid)

```